Scaling through abstractions – high-performance vectorial wave simulations for seismic inversion with Devito

### 2020-06-01 response.md

Dear Editor,

First of all, we would like to thank the reviewers for their work and constructive feedback. Below, we address all of their questions and observations.

# Review of pap429s2 by Reviewer 1
Detailed Comments for Authors
    This is an interesting paper, well-written, logically ordered and shows interesting results. I only have small questions having to do with scaling results and validation.
Comments for Revision
    If you run your model with a variable number of cores, how is the weak and strong scaling? Can you comment on that? Further, you mention that you validate performance against a wave propagator - how about validation of simulation quality? How is that done?
Scored Review Questions ACCEPT (4)

At the time of the experimentation, we ran into two main issues:
1. Limited resources on Azure, including the access to an Infiniband-enabled system.
2. Most of the MPI-level optimizations that are today in Devito were still in a tuning and consolidation phase.

We subsequently managed to work around point one above; we have published some scaling experiments in [..., page ..., figure ...], although they still don't exploit the latest advancements in Devito v4.2 (e.g., computation/communication overlap). What we want to remark is that our manuscript centers on the development of HPC production-grade wave propagators from a high-level language, and while performance is one of the cornerstones of our work, it was not the objective of this paper to show how powerful Devito is in an extreme scaling setting. As -- we are sure -- you have understood, this is completely sensible for a venue such as SC. We could have added scaling plots showing what, with the latest Devito v4.2, appears to be a remarkable parallel efficiency, but we feel that this would have been premature, and our philosophy is not to publish any results that has not been adequately validated and reproduced over multiple architectures. This attitude is at the hearth of the success of Devito, and it's not our intention to undermine it for a mere publication.

About validation of simulation quality -- [TODO]

# Review of pap429s2 by Reviewer 2
Detailed Comments for Authors
    This paper presents a methodology to use abstractions and compiler technology effectively in the field of exploration seismology. It is a very nicely written paper that puts down all the motivation and technical details quite well. However, it is not very suitable for SC. Parallelization is pretty straightforward, uniform decomposition of relatively straightforward operation. The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization, and therefore not very interesting to this community. The parallelism applied so far is straight up MPI which is a bit behind times these days.
Comments for Revision
    I don't think any revision will change the nature of the work, which, as I mentioned in comments section is nicely done but not best suited for SC.
Scored Review Questions REJECT (1)

With all due respect, we strongly disagree with all the critiques moved to the manuscript. Let's go through them one at a time.

* However, it is not very suitable for SC

As partly explained in our reply to R1, this was conceived as an application oriented paper exploiting high-level abstractions and compilation technology. Computational science is evolving rapidly, with production-level DSL-based systems replacing unmaintainable legacy codes, and our work is just yet another small brick in this direction. Unlike the majority of DSL-based tools that often get published, our work builds on and contributes to an established framework for seismic imaging, so we believe that this should be of strong interest for the SC community.

*  Parallelization is pretty straightforward, uniform decomposition of relatively straightforward operation.

First, we observe that this is not just manual implementation of some MPI-based domain decomposition for some pre-existing code. This is rather _generation_ of MPI parallel code from a purely mathematical description of solvers for partial differential equations. The software engineering required is considerable, especially for an open source project such as Devito that was born in academia. We're talking about thousands of lines of carefully engineered code, plus tests and examples, which allow users to just run `mpirun -n X python ...` to run over MPI with virtually no changes to code. Also, as we explained in the paper, it's not grid decomposition only -- sparse functions, distributed numpy arrays, compiler-level optimazion for effective halo exchange -- these are just a few examples of features that are extremely far from being regarded as "pretty straightforward".

* The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization

This is simply an error. A high arithmetic intensity does not imply simple optimization. In fact, it's entirely the opposite, and this returns over and over again in computational science. For example, with high-order finite (spectral) elements -- you may get an extremely good OI, but if you're not using tensor-product basis functions and suitable techniques such as sum-factorization, yes, you'll perform loads of flops and obtain a beautiful GFlops/s performance, but unfortunately a huge fraction of those flops will be totally useless, as redundant. The same story happens with our TTI. As we explain in the paper, without having captured all those cross-iteration common sub-expressions,  we would have obtained remarkable GFlops/s performance, but terrible GPoints/s. And capturing those redundancies is extremely difficult -- in fact, this is one of the reasons several companies are today evaluating Devito: machines (compilers) can be better than humans at detecting redundancies, and definitely way faster than them. In the paper we spend a few paragraphs describing how this is achieved. We think this has potentially great interest for the SC community.

* The parallelism applied so far is straight up MPI which is a bit behind times these days.

As far as we know, in 2020 MPI is still the most widely used approach for domain decomposition.

# Review of pap429s2 by Reviewer 3
Detailed Comments for Authors
    The authors in the paper describe their Devito, open-source Python project targeted in the area of seismic exploration. The idea of this particular piece of work is the ease of use through Devito to address, in a Domain Specific Language, creating code for seismic exploration problems which due to being computational intensive often require the use of distributed parallel systems. This is built upon MPI but appears to be hidden from the end user. It was a little confusing but the MPI parallelism appears to essentially be a domain decomposition style parallelism. There was some mention of threads, in particular there seems to be several compilation passes that abstracts out the parallelism which can include SIMD, shared-memory and MPI. Only the MPI is detailed. Through Devito, the authors demonstrate its utility on two problems; (1) Reverse Time Migration (RTM) for a pseudo-acoustic wave problem and (2) an elastic wave problem. With the RTM, they consider titled-transverse isotropic (TTI) wave equation formulation. In RTM, the forward problem is calculated and then the adjoint problem is solve backwards in time to get the imaging. RTM tends to require significant memory and/or fast access to storage of the forward problem during solves for the adjoint problem. Devito is used to set these problems up in VM (Virtual Machines) to be run in a Cloud (Azure) on 32 nodes. If this reviewer understands correctly, Devito is set up to calculate the FLOPs count which is how the authors get the performance. One question that arises, in the performance is how does Devito account for FMA (Fused Multiple Adds)? A comparison was made with an open source hand-code propagator "fdelmode" for the elastic wave equation which seems to be on a 2D problem. The result of this comparison are "essentially identical". A table comparison might be helpful. Scalability performance would also be interesting.
    It appears much of the performance result are provided in the references so it is unclear as to what the purpose of this paper is for an SC audience. For actual seismic surveys, the imaging (interpretation) part is only 10% of the survey. Of course, one does us the RTM in developing the velocity model which adds to complexity in a way. Ease of use coupled with excellent scalable performance while important doesn't address the real "industrial" problems, it helps a little.
Comments for Revision
    Perhaps restricting the discussion of both ease of use of Devito, with a comparison including scalability on one problem, say the 3D TTI problem against a known method and doing this in a cloud setting up to a significant number of nodes would make this a more appropriate paper for the SC audience. This might be too tall an order here. Dropping the discussion on the 3D elastic wave and including a table of comparison with the fdelmode for the 2D elastic wave problem or even just going through scalability comparison on the TTI problem against another code might demonstrate the value of Devito better.
Scored Review Questions WEAK REJECT (2)

# Review of pap429s2 by Reviewer 4 
Detailed Comments for Authors
    This manuscript describes optimizations for Devito, a symbolic DSL designed for geophysics, to extend its use for industrial scale modeling of seismic inversion problems.
    Strengths:
        added support for vector and tensor objects for staggered-grid finite-differences
        demonstrated solving large-scale inverse problems (requiring large memory and computation)
        demonstrated its use on Cloud-based HPC
        paper is well-written
    Weaknesses:
        One issue is that I cannot tell how much of this work is new
        contributions. The paper reads a lot like an overview of previous work in many parts. The parts that I am sure are new work (e.g., adding support for vector and tensors), I am unclear of the amount of effort or innovation involved.
        I like this paper and the work (the ease-of-use of Devito is impressive), but I think for SC we would need more information on its scalability and performance.
Comments for Revision
        Please address the weaknesses above
        Is section III describing new work? It is unclear to me if this is an overview or new work for this paper…
        Is the parallelism just via MPI? What are the possibilities for this code in terms of heterogeneous clusters? Will it be able to take advantage of GPUs?
    Minor issues:
    (1) section 3, 1st paragraph: "must to be" => "must be"
    (2) section 1, 3rd paragraph: make an itemized list?
Scored Review Questions WEAK REJECT (2)

### 2020-06-01 response.md

Dear Editor,

First of all, we would like to thank the reviewers for their work and constructive feedback. Below, we address all of their questions and observations.

## Review of pap429s2 by Reviewer 1
Detailed Comments for Authors
    This is an interesting paper, well-written, logically ordered and shows interesting results. I only have small questions having to do with scaling results and validation.
Comments for Revision
    If you run your model with a variable number of cores, how is the weak and strong scaling? Can you comment on that? Further, you mention that you validate performance against a wave propagator - how about validation of simulation quality? How is that done?
Scored Review Questions ACCEPT (4)

### Reply

@@

At the time of the experimentation, we ran into two main issues:
1. Limited resources on Azure, including the access to an Infiniband-enabled system.
2. Most of the MPI-level optimizations that are today in Devito were still in a tuning and consolidation phase.

We subsequently managed to work around point one above; we have published some scaling experiments in [..., page ..., figure ...], although they still don't exploit the latest advancements in Devito v4.2 (e.g., computation/communication overlap). What we want to remark is that our manuscript centers on the development of HPC production-grade wave propagators from a high-level language, and while performance is one of the cornerstones of our work, it was not the objective of this paper to show how powerful Devito is in an extreme scaling setting. As -- we are sure -- you have understood, this is completely sensible for a venue such as SC. We could have added scaling plots showing what, with the latest Devito v4.2, appears to be a remarkable parallel efficiency, but we feel that this would have been premature, and our philosophy is not to publish any results that has not been adequately validated and reproduced over multiple architectures. This attitude is at the hearth of the success of Devito, and it's not our intention to undermine it for a mere publication.

About validation of simulation quality -- [TODO]

## Review of pap429s2 by Reviewer 2
Detailed Comments for Authors
    This paper presents a methodology to use abstractions and compiler technology effectively in the field of exploration seismology. It is a very nicely written paper that puts down all the motivation and technical details quite well. However, it is not very suitable for SC. Parallelization is pretty straightforward, uniform decomposition of relatively straightforward operation. The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization, and therefore not very interesting to this community. The parallelism applied so far is straight up MPI which is a bit behind times these days.
Comments for Revision
    I don't think any revision will change the nature of the work, which, as I mentioned in comments section is nicely done but not best suited for SC.
Scored Review Questions REJECT (1)

### Reply

@@

This is simply an error. A high arithmetic intensity does not imply simple optimization. In fact, it's entirely the opposite, and this returns over and over again in computational science. For example, with high-order finite (spectral) elements -- you may get an extremely good OI, but if you're not using tensor-product basis functions and suitable techniques such as sum-factorization, yes, you'll perform loads of flops and obtain a beautiful GFlops/s performance, but unfortunately a huge fraction of those flops will be totally useless, as redundant. The same story happens with our TTI. As we explain in the paper, without having captured all those cross-iteration common sub-expressions,  we would have obtained remarkable GFlops/s performance, but terrible GPoints/s. And capturing those redundancies is extremely difficult -- in fact, this is one of the reasons several companies are today evaluating Devito: machines (compilers) can be better than humans at detecting redundancies, and definitely way faster than them. In the paper we spend a few paragraphs describing how this is achieved. We think this has potentially great interest for the SC community.

* The parallelism applied so far is straight up MPI which is a bit behind times these days.

As far as we know, in 2020 MPI is still the most widely used approach for domain decomposition.

## Review of pap429s2 by Reviewer 3
Detailed Comments for Authors
    The authors in the paper describe their Devito, open-source Python project targeted in the area of seismic exploration. The idea of this particular piece of work is the ease of use through Devito to address, in a Domain Specific Language, creating code for seismic exploration problems which due to being computational intensive often require the use of distributed parallel systems. This is built upon MPI but appears to be hidden from the end user. It was a little confusing but the MPI parallelism appears to essentially be a domain decomposition style parallelism. There was some mention of threads, in particular there seems to be several compilation passes that abstracts out the parallelism which can include SIMD, shared-memory and MPI. Only the MPI is detailed. Through Devito, the authors demonstrate its utility on two problems; (1) Reverse Time Migration (RTM) for a pseudo-acoustic wave problem and (2) an elastic wave problem. With the RTM, they consider titled-transverse isotropic (TTI) wave equation formulation. In RTM, the forward problem is calculated and then the adjoint problem is solve backwards in time to get the imaging. RTM tends to require significant memory and/or fast access to storage of the forward problem during solves for the adjoint problem. Devito is used to set these problems up in VM (Virtual Machines) to be run in a Cloud (Azure) on 32 nodes. If this reviewer understands correctly, Devito is set up to calculate the FLOPs count which is how the authors get the performance. One question that arises, in the performance is how does Devito account for FMA (Fused Multiple Adds)? A comparison was made with an open source hand-code propagator "fdelmode" for the elastic wave equation which seems to be on a 2D problem. The result of this comparison are "essentially identical". A table comparison might be helpful. Scalability performance would also be interesting.
    It appears much of the performance result are provided in the references so it is unclear as to what the purpose of this paper is for an SC audience. For actual seismic surveys, the imaging (interpretation) part is only 10% of the survey. Of course, one does us the RTM in developing the velocity model which adds to complexity in a way. Ease of use coupled with excellent scalable performance while important doesn't address the real "industrial" problems, it helps a little.
Comments for Revision
    Perhaps restricting the discussion of both ease of use of Devito, with a comparison including scalability on one problem, say the 3D TTI problem against a known method and doing this in a cloud setting up to a significant number of nodes would make this a more appropriate paper for the SC audience. This might be too tall an order here. Dropping the discussion on the 3D elastic wave and including a table of comparison with the fdelmode for the 2D elastic wave problem or even just going through scalability comparison on the TTI problem against another code might demonstrate the value of Devito better.
Scored Review Questions WEAK REJECT (2)

### Reply

...

## Review of pap429s2 by Reviewer 4 
Detailed Comments for Authors
    This manuscript describes optimizations for Devito, a symbolic DSL designed for geophysics, to extend its use for industrial scale modeling of seismic inversion problems.
    Strengths:
        added support for vector and tensor objects for staggered-grid finite-differences
        demonstrated solving large-scale inverse problems (requiring large memory and computation)
        demonstrated its use on Cloud-based HPC
        paper is well-written
    Weaknesses:
        One issue is that I cannot tell how much of this work is new
        contributions. The paper reads a lot like an overview of previous work in many parts. The parts that I am sure are new work (e.g., adding support for vector and tensors), I am unclear of the amount of effort or innovation involved.
        I like this paper and the work (the ease-of-use of Devito is impressive), but I think for SC we would need more information on its scalability and performance.
Comments for Revision
        Please address the weaknesses above
        Is section III describing new work? It is unclear to me if this is an overview or new work for this paper…
        Is the parallelism just via MPI? What are the possibilities for this code in terms of heterogeneous clusters? Will it be able to take advantage of GPUs?
    Minor issues:
    (1) section 3, 1st paragraph: "must to be" => "must be"
    (2) section 1, 3rd paragraph: make an itemized list?
Scored Review Questions WEAK REJECT (2)

### Reply

...

### 2020-06-01 response.md

### Reply

* One question that arises, in the performance is how does Devito account for FMA (Fused Multiple Adds)?

The operations determining the performance of a code are the source level ones, while FMA is a lower level concept (ie, a scalar FMA would account for two operations). This is not Devito, but rather then model adopted everywhere, otherwise two performance values from two different architectures, one with FMAs the other without, would not even be comparable.

* Scalability performance would also be interesting.

[TODO] Reuse part of replies above

*     It appears much of the performance result are provided in the references so it is unclear as to what the purpose of this paper is for an SC audience. For actual seismic surveys, the imaging (interpretation) part is only 10% of the survey. Of course, one does us the RTM in developing the velocity model which adds to complexity in a way. Ease of use coupled with excellent scalable performance while important doesn't address the real "industrial" problems, it helps a little.

[TODO] Felix, Gerard, Mathias...

### 2020-06-01 paper.tex
rhodrin mods + tidying.

@@

This paper is organized as follows: first, we provide a brief overview of
Devito and its symbolic API and present the distributed memory
implementation that allows large-scale modelling and inversion by means of
domain decomposition. We then provide a brief comparison with a state of
the art hand-coded wave propagator to validate the performance previously
benchmarked with the roofline model (\cite{patterson, devito-compiler,
devito-api, louboutin2016ppf}).  Before concluding, results from the
Cloud-based experiments discussed above are presented, highlighting the
vectorial and tensorial capabilities of Devito.

@@

Devito \cite{devito-api} provides a functional language built on top of
\texttt{SymPy} \cite{sympy} to symbolically express PDEs at a
mathematical level and implements automatic discretization with the
finite-difference method. The language is by design flexible enough to
express other types of non finite-difference operators, such as
interpolation and tensor contractions, that are inherent to
measurement-based inverse problems. Several additional features are
supported, among which are staggered grids, sub-domains, and stencils with
custom coefficients. The last major building block of a solid PDE solver
are the boundary conditions which for finite-difference methods are
notoriously diverse and often complicated. The system is, however,
sufficiently general to express them through a composition of core
mechanisms. For example, free surface and perfectly-matched layers
(PMLs) boundary conditions can be expressed as equations -- just like
any other PDE equations -- over a suitable sub-domain.

@@

In this section we illustrate the \devito language by demonstrating
an implementation of the acoustic wave-equation in isotropic media
%
\begin{equation}
\begin{cases}
 m \frac{d^2 u(t, x)}{dt^2} - \Delta u(t, x) = \delta(xs) q(t) \\
 u(0, .) = \frac{d u(t, x)}{dt}(0, .) = 0 \\
 d(t, .) = u(t, xr).
 \end{cases}
\label{acou}
\end{equation}
%
 The core of the Devito
symbolic API consists of three classes:

@@

A \texttt{Grid} represents a discretized finite n-dimensional space and
is created as follows

\begin{python}[label=grid, caption=Grid creation]
from devito import Grid
grid = Grid(shape=(nx, ny, nz),
            extent=(ext_x, ext_y, ext_z),
            origin=(o_x, o_y, o_z))
\end{python}
\noindent
where \texttt{(nx,\phantom{\ }ny,\phantom{\ }nz)} are the number of grid
points in each direction,
\texttt{(ext\_x,\phantom{\ }ext\_y,\phantom{\ }ext\_z)} is the physical
extent of the domain in physical units (i.e \texttt{m}) and
\texttt{(o\_x,\phantom{\ }o\_y,\phantom{\ }o\_z)} is the origin of the
domain in the same physical units. The object \texttt{grid} contains all
the information related to the discretization such as the grid spacing.
We use \texttt{grid} to create the symbolic objects that will be used to
express the wave-equation. First, we define a spatially varying model
parameter \texttt{m} and a time-space varying field \texttt{u}

@@

The main complexity of the TTI wave-equation is that the rotation of the
symmetry axis of the physics leads to rotated second-order finite-difference
stencils. In order to ensure numerical stability, these rotated
finite-difference operators are designed to be self-adjoint (c.f.
\cite{zhang2011, duveneck}). For example, we define the rotated second order
derivative with respect to $x$ as:
%
\begin{equation}
\begin{aligned}
  G_{\bar{x}\bar{x}} &= D_{\bar{x}}^T D_{\bar{x}} \\
  D_{\bar{x}} &= \cos(\mathbf{\theta})\cos(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}x} + \cos(\mathbf{\theta})\sin(\mathbf{\phi})\frac{\mathrm{d}}{\mathrm{d}y} - \sin(\mathbf{\theta})\frac{\mathrm{d}}{\mathrm{d}z}.
\end{aligned}
\label{rot}
\end{equation}
%

@@

The isotropic elastic wave-equation, parametrized by the so-called Lamé parameters
$\lambda, \mu$ and the density $\rho$ reads:
%
\begin{equation}
\begin{aligned}
&\frac{1}{\rho}\frac{dv}{dt} = \nabla . \tau \\
&\frac{d \tau}{dt} = \lambda \mathrm{tr}(\nabla v) \mathbf{I}  + \mu (\nabla v + (\nabla v)^T)
\end{aligned}
\label{elas1}
\end{equation}
%
\noindent
 where $v$ is a vector valued function with one component per cartesian
direction:
%
\begin{equation}
\begin{split}
v =  \begin{bmatrix} v_x(t, x, y) \\ v_y(t, x, y)) \end{bmatrix}
\end{split}
\label{partvel}
\end{equation}
%
 and the stress $\tau$ is a symmetric second-order tensor-valued
function:
%
\begin{equation}
\begin{aligned}
    \tau = \begin{bmatrix}\tau_{xx}(t, x, y) & \tau_{xy}(t, x, y)\\\tau_{xy}(t, x, y) & \tau_{yy}(t, x, y)\end{bmatrix}.
\end{aligned}
\label{stress}
\end{equation}
%
 The discretization of \eqref{elas1} and \eqref{stress} requires five
equations in two dimensions (two equations for the particle velocity and
three for the stress) and nine equations in three dimensions (three for the particle velocity
and six for the stress). However, the mathematical
definition only require two coupled vector/tensor-valued equations for
any number of dimensions.

@@

Earlier performance benchmarks mainly focused on roofline model analysis.
In this study, for completeness, the runtime of \devito is therefore
compared to that of the open source hand-coded propagator
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc}. This propagator,
described in \cite{thorbecke}, is a state of the art elastic kernel
(Equation~\ref{elas1}) and the comparisons presented here were carried out
in collaboration with its author. To ensure a fair
comparison, we ensured that the
physical and computational settings were identical. The settings were as
follows:

@@

The runtimes observed for this problem were essentially identical,
showing less than a one percent of difference. Such similar runtimes were
obtained with both the Intel and GNU compilers and the
experiment was performed with both fourth and sixth order discretizations.
Kernels were executed five times each and the
runtimes observed were consistently very similar. This comparison
illustrates the performance achieved with \devito is at least on par with
hand-coded propagators. Considering we do not take advantage of the Devito
compilers full capabilities in two dimensional cases, we are confident
that the code generated will be at least on par with the hand-coded version
for three dimensional problems and this comparison will be part of our future
work.

@@

Transitioning from academic toy problems, such as the two-dimensional
acoustic wave-equation, to real-world applications can be challenging,
particularly if this transition is carried out as an afterthought.
Owing to the fundamental design principles of \devito such scaling,
however, becomes trivial. In this work we demonstrated
the high-level interface provided by \devito not only for simple scalar
equations but also for coupled PDEs. This interface allows, in a simple, concise
and consistent manner, the expression of all-kinds of non-trivial
differential operators. Next, and most
importantly, we demonstrated that the compiler enables large-scale modelling with
state-of-the art computational performance and programming paradigm. The
single-node performance is on par with state of the art hand-coded models, but
packaged with this performance comes
the flexibility of the symbolic interface and multi-node parallelism, which is
integrated in the compiler and interface in a accessible way. Finally,
we demonstrated that our abstractions provide the necessary portability
to enable both on-premise and Cloud based HPC.

>>>>

\textcolor{red}{It should be emphasized that unlike in,
e.g., earthquake simulation models \cite{fu201718} (more?), where
a single shot is computed, wave-equation based
geophysical inversion not only requires
fast and accurate solvers but also adavnced algorithms for
non-convex optimization of large scale constrained inverse problems. The 
high-level interface provided by \devito enables research and developement for these 
specific problem in a framework that scales to industry size problem 
automatically through its compiler. \devito has been used previously
for the solution of the wave equation within imaging frameworks including
FWI \cite{witte2018alf}, compressived least-square RTM \cite{witte2018cls} and
time-domain Wavefield Reconstruction Inversion (WRI)~\cite{rizzuti2020EAGEtwri, 
louboutin2020SEGtwri}, the later of which takles the cycle-skipping issue
present in seismic inversion.}

@@

This paper is organized as follows: first, we provide a brief overview of
Devito and its symbolic API and present the distributed memory
implementation that allows large-scale modelling and inversion by means of
domain decomposition. We then provide a brief comparison with a state of
the art hand-coded wave propagator to validate the performance previously
benchmarked with the roofline model (\cite{patterson, devito-compiler,
devito-api, louboutin2016ppf}). \textcolor{red}{Before concluding, results from the
Cloud-based experiments introduced above are presented}.

@@

Devito \cite{devito-api} provides a functional language built on top of
\texttt{SymPy} \cite{sympy} to symbolically express PDEs at a
mathematical level and implements automatic discretization with the
finite-difference method. The language is by design flexible enough to
express other types of non finite-difference operators, such as
interpolation and tensor contractions, that are inherent to
measurement-based inverse problems. Several additional features are
supported, among which are staggered grids, sub-domains, and stencils with
custom coefficients. The last major building block of a solid PDE solver
are the boundary conditions which for finite-difference methods are
notoriously diverse and often complicated. The system is, however,
sufficiently general to express them through a composition of core
mechanisms. For example, free surface and perfectly-matched \textcolor{red}{layer
(PML) conditions} can be expressed as equations -- just like
any other PDE equations -- over a suitable sub-domain.

@@

In this section we illustrate the \devito language by demonstrating
an implementation of the acoustic wave-equation in isotropic media
%
\begin{equation}
\begin{cases}
 m \frac{\partial^2 u(t, x)}{\partial t^2} - \Delta u(t, x) = \delta(xs) q(t) \\
 u(0, .) = \frac{\partial u(t, x)}{\partial t}(0, .) = 0 \\
 d(t, .) = u(t, xr).
 \end{cases}
\label{acou}
\end{equation}
%
\textcolor{red}{FIXME: Define the terms appearing in the above equation.}
 The core of the Devito
symbolic API consists of three classes:

@@

A \texttt{Grid} represents a discretized finite $n$-dimensional space and
is created as follows

\begin{python}[label=grid, caption=Grid creation]
from devito import Grid
grid = Grid(shape=(nx, ny, nz),
            extent=(ext_x, ext_y, ext_z),
            origin=(o_x, o_y, o_z))
\end{python}
\noindent
where \texttt{(nx,\phantom{\ }ny,\phantom{\ }nz)} are the number of grid
points in each direction,
\texttt{(ext\_x,\phantom{\ }ext\_y,\phantom{\ }ext\_z)} is the
extent of the domain in physical units (i.e \texttt{m}) and
\texttt{(o\_x,\phantom{\ }o\_y,\phantom{\ }o\_z)} is the origin of the
domain in the same physical units. \textcolor{red}{The \texttt{grid} object contains all
information pertinent to the discretization such as the grid spacing}.
We use \texttt{grid} to create the symbolic objects that will be used to
express the wave-equation. First, we define a spatially varying model
parameter \texttt{m} and a time-space varying field \texttt{u}

@@

The main complexity of the TTI wave-equation is that the rotation of the
symmetry axis of the physics leads to rotated second-order finite-difference
stencils. In order to ensure numerical stability, these rotated
finite-difference operators are designed to be self-adjoint (c.f.
\cite{zhang2011, duveneck}). For example, we define the rotated second order
derivative with respect to $x$ as:
%
\begin{equation}
\begin{aligned}
  G_{\bar{x}\bar{x}} &= D_{\bar{x}}^T D_{\bar{x}} \\
  D_{\bar{x}} &= \cos(\mathbf{\theta})\cos(\mathbf{\phi})\frac{\partial}{\partial x} + \cos(\mathbf{\theta})\sin(\mathbf{\phi})\frac{\partial}{\partial y} - \sin(\mathbf{\theta})\frac{\partial}{\partial z}.
\end{aligned}
\label{rot}
\end{equation}
%

@@

The isotropic elastic wave-equation, parametrized by the so-called Lamé parameters
$\lambda, \mu$ and the density $\rho$ reads:
%
\begin{equation}
\begin{aligned}
&\frac{1}{\rho}\frac{\partial v}{\partial t} = \nabla . \tau \\
&\frac{\partial \tau}{\partial t} = \lambda \mathrm{tr}(\nabla v) \mathbf{I}  + \mu (\nabla v + (\nabla v)^T)
\end{aligned}
\label{elas1}
\end{equation}
%
\noindent
 where $v$ is a vector valued function with one component per cartesian
direction:
%
\begin{equation}
\begin{split}
v =  \begin{bmatrix} v_x(t, x, y) \\ v_y(t, x, y)) \end{bmatrix}
\end{split}
\label{partvel}
\end{equation}
%
 and the stress $\tau$ is a symmetric second-order tensor-valued
function:
%
\begin{equation}
\begin{aligned}
    \tau = \begin{bmatrix}\tau_{xx}(t, x, y) & \tau_{xy}(t, x, y)\\\tau_{xy}(t, x, y) & \tau_{yy}(t, x, y)\end{bmatrix}.
\end{aligned}
\label{stress}
\end{equation}
%
 Note that the discretization of \eqref{elas1} and \eqref{stress} requires five
scalar equations in two dimensions (two for the particle velocity and
three for stress) and nine equations in three dimensions (three for particle velocity
and six for stress). However, the mathematical
definition only require two coupled vector/tensor-valued equations for
any number of dimensions.

@@

Earlier performance benchmarks mainly focused on roofline model analysis
\textcolor{red}{[TODO: Add reference]}.
In this study, for completeness, the runtime of \devito is therefore
compared to that of the open source hand-coded propagator
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc}. This propagator,
described in \cite{thorbecke}, is a state of the art elastic kernel
(Equation~\ref{elas1}) and the comparisons presented here were carried out
in collaboration with its author.
\textcolor{red}{It should be emphasized that direct model comparison in
the area of exploration geophysics is challenging. The majority of production
codes are either proprioty or are simply not publicly available. Further,
\devito should not be compared against pure stencil frameworks,
such as YASK \cite{yask}, since they are unable to compute the sparse operations
required for source injection and receiver interpolation (and are therefore not
suitable for use in measurement based inversion problems). The comparison carried out
here is conducted in two dimensions. Whilst large scale 3D scaling comparisons are theoretically
possible, compute time on the Cloud is expensive. Such a comparison is not feasible at this
time and will therefore be left for future work.}

To ensure a fair comparison, we ensured that the
physical and computational settings were identical. The settings were as
follows:

@@

The runtimes observed for this problem were essentially identical,
showing less than a one percent of difference. Such similar runtimes were
obtained with both the Intel and GNU compilers and the
experiment was performed with both fourth and sixth order discretizations.
Kernels were executed five times each and the
runtimes observed were consistently very similar. This comparison
illustrates the performance achieved with \devito is at least on par with
hand-coded propagators. Considering we do not take advantage of the Devito
compilers full capabilities in \textcolor{red}{the two dimensional case}, we are confident
that the code generated will be at least on par with the hand-coded version
for three dimensional problems and this comparison will be part of our future
work.

@@

Transitioning from academic toy problems, such as the two-dimensional
acoustic wave-equation, to real-world applications can be challenging,
particularly if this transition is carried out as an afterthought.
Owing to the fundamental design principles of \devito such scaling,
however, becomes trivial. In this work we demonstrated
the high-level interface provided by \devito not only for simple scalar
equations but also for coupled PDEs. This interface allows, in a simple, concise
and consistent manner, the expression of all-kinds of non-trivial
differential operators. Next, and most
importantly, \textcolor{red}{it was} demonstrated that the compiler enables large-scale modelling with
state-of-the art computational performance and programming paradigm. The
single-node performance is on par with state of the art hand-coded models, but
packaged with this performance comes
the flexibility of the symbolic interface and multi-node parallelism, which is
integrated in the compiler and interface in a accessible way. Finally,
we demonstrated that our abstractions provide the necessary portability
to enable both on-premise and Cloud based HPC.
